\Ace: An Extensible Static Type System Inside a Python

### 2014-03-26 ace-oopsla14/string-appendix.tex
Content finished; starting cleanup.

TODO-nrf spell check.

TODO-nrf Include functions (app, abs)?

TODO-nrf Change icheck in eval rule to Python.

TODO-nrf Is it ok to make the proofs single-column? If not clean up overflows.

TODO-nrf Make sure failure of unicity isn't causing any problems and write down lemmas about
forms (e.g. $e \in \lang{r} \implies e = \strf{s}$ etc).

@@

\inferline{E-concat}
{e_1 + e_2 : \strin[r_1+r_2] \ireduces \istrf{e_1} + \istrf{e_2}}
{ }

@@

\inferline{E-convertval}
{ [\strin[r']](e) : \strin[r'] \reduces [\strin[r']](e') }
{ e : \strin[r] \reduces e' }

\inferline{E-convert}
{ [\strin[r']](e) : \strin[r'] \ireduces \istrf{e} }
{ e \val }

\inferline{E-dconvertval (optional) \footnote{See definition ???} }
{\dconvert{\strin[r']}{e} : \strin[r'] \reduces {\dconvert{\strin[r']}{e'}}}
{e:\strin[r] \reduces e'}

\inferline{E-dconvert (optional) \footnote{See definition ???} }
{\dconvert{\strin[r']}{e} : \strin[r'] \ireduces \ifilter{r'}{\istrf{e}}}
{e \val }

\end{tabular}
\caption{Operational semantics for $\lcs$}
\label{fig:lcsEval}
\end{center}
\end{figure}
\twocolumn

@@

The type safety proof relies upon some established results about regular languages.
The important definitions and theorems follow.

\subsection{Properties of Regular Languages}
 % copy from thesis

@@

Before presenting progress and preservation theorems, we establish some definitions
and lemmas necessary for the proofs.

\begin{defn}[Types of internal values]
The types of internal values are defined as follows:
\begin{itemize}
  \item If $e=\istrf{s}$ then $e:\istr$.
  \item If $e=\ifilter(r, \istrf{s})$ then $e:\istr$.
  \item If $e=\istrf{s_1}+\istrf{s_2}$ then $e:\istr$.
\end{itemize}
\end{defn}

Definition 2 is an additional assumption about
the internal language. To summarize, we assume an internal language containing 
strings together with operations for string concatentation and filtering. We
expect closure over strings for both operations.
Recall that $\ifilter$ is only needed for dynamic casts, which may be removed without
descreasing the expressivity or even usability of the language.

@@

\begin{thm}[Preservation]
  Let $T$ be a type in $\lcs$ and $h(T)=\sigma$ the corresponding type in the internal language.
  For all terms $e$:
  \begin{itemize}
    \item If $e:T$ and $e:T \ireduces i$ then $i : \sigma$ for some $\sigma$.
    \item If $e:T$ and $e:T \reduces e'$ then $e':T$.
  \end{itemize}
\end{thm}
\begin{proof}
The proof is a straightforward induction on the derivation of the combined evaluation relation.
\begin{itemize}[label=$ $,itemsep=1ex]
  \item \textbf{E-Ival, E-Ifilterval}. Both cases hold since the terms at hand are not $\lcs$ terms.
  \item \textbf{E-Stringval, E-strval}. Both cases hold since no reduction is possible. 
  \item \textbf{E-String}. By the definition of typing for internal terms, $\istrf{e} : \istr$. It suffices to show that $h(\str) = \istr$, which follows from the definition of $h$.
  \item \textbf{E-String\_in}. By the definition of typing for internal terms, $\istrf{e} : \istr$. It suffices to show that $h(\strin[r]) = \istr$, which follows from the definition of $h$ for arbitrary $r$.
  \item \textbf{E-concat}. By the definition of typing for internal terms, $\istrf{e_1} + \istrf{e_2} : \istr$. So it suffices to show that $h(\strin[r_1+r_2]) = \istr$, which follows from the definition of $h$ for arbitrary $r_1$, $r_2$.
  \item \textbf{E-Filterval}. We have that $\filter{\strin[r']}{e}:\strin[r\backslash r' + \emptyset]$. By inversion of T-Filter, $e:\strin[r]$.
    By T-Equiv-string\_in (which is bidirectional), $e \in \lang{r}$.
    By theorem 2, ${\tt rl\_filter}(r,e) \in \lang{r\backslash r' + \emptyset}$.
    By T-Equiv-string\_in,  $e \in \lang{r}$ and ${\tt rl\_filter}(r,e) \in \lang{r\backslash r' + \emptyset}$ implies ${\tt rl\_filter}(r,e):\strin[r\backslash r' + \emptyset]$.
  \item \textbf{E-Filter}. By inversion, $e : \strin[r'] \reduces e'$ so by the induction $e':\strin[r']$.
    Therefore, ${\tt filter}(\strin[r], e') : \strin[r \backslash r' + \emptyset]$ by T-Filter.
  \item \textbf{E-Convertval}. It suffices to show that $h(\strin[r]) = \istr$, which is true by definition.
  \item \textbf{E-Convert}. By inversion and induction, $e' : \strin[r]$. We know that $[\strin[r']](e) : \strin[r']$, so by inversion of T-Convert $\lang{r'} \subseteq \lang{r}$.
    It follows that $[\strin[r']](e') : \strin[r']$.
  \item \textbf{E-DConvert}. By inversion, $e : \strin[r'] \reduces e'$. By the induction hypothesis, $e' : \strin[r']$; therefore, by T-Dconvert, $\dconvert[\strin[r]](e):\strin[r]$.
  \item \textbf{E-DConvertval}. By the definition of typing for internal terms, $\ifilter(r,\istrf{e}) : \istr$. It suffices to show that $h(\strin[r_1+r_2]) = \istr$, which follows from the definition of $h$.
\end{itemize}
\end{proof}

\begin{thm}[Progress]
  If $e:T$ then $e:T \reduces^* e' \ireduces^* i$ where $i \ival$.
\end{thm}
\begin{proof}
By induction on the derivation of $e:T$.
\begin{itemize}[label=$ $,itemsep=1ex]
  \item \textbf{T-Equiv-Include}. Since $e \in \lang{r}$, $e = \strf{s}$ for some $s$; therefore, $e \val$ by E-Strinval.
  \item \textbf{T-Minimum}. 
\end{itemize}
\end{proof}

@@

>>>>

@@

\inferline{E-concatval}
{e_1 + e_2 : \strin[r_1+r_2] \ireduces \istrf{e_1} + \istrf{e_2}}
{e_1 \val \\ e_2 \val }

\inferline{E-concatL}
{e_1 + e_2 : \strin[r_1+r_2] \reduces {e_1}' + {e_2}}
{ e_1 : \strin[r_1] \reduces {e_1}'}

\inferline{E-concatR}
{e_1 + e_2 : \strin[r_1+r_2] \reduces {e_1} + {e_2}'}
{ e_2 : \strin[r_2] \reduces {e_2}'}

@@

\inferline{E-convertval}
{ [\strin[r']](e) : \strin[r'] \ireduces \istrf{e} }
{ e \val }

\inferline{E-convert}
{ [\strin[r']](e) : \strin[r'] \reduces [\strin[r']](e') }
{ e : \strin[r] \reduces e' }

\inferline{E-dconvertval (optional) }
{\dconvert{\strin[r']}{e} : \strin[r'] \reduces {\dconvert{\strin[r']}{e'}}}
{e:\strin[r] \reduces e'}

\inferline{E-dconvert (optional) }
{\dconvert{\strin[r']}{e} : \strin[r'] \ireduces \ifilter{r'}{\istrf{e}}}
{e \val }

\end{tabular}
\caption{SOS rules for $\lcs$}
\label{fig:lcsEval}
\end{center}
\end{figure}

\twocolumn

The type safety proof relies upon some assumptions about the type system and
dynamics of the internal language, as well as some properties of regular
languages.

There must exist a translation from $\lcs$ types to the types of the internal
language. For the remainder of this paper, we call the type translation function
$h$.

@@

Additionally, we assume that the internal language contains an implementation of strings,
together with operations for concatenation and filtering by regular expression.

\begin{defn}[Types of internal values]
  Let $\istrf{s}$ range over string literals and $r$ over regular expressions.
  Internal values are typed as follows:
\begin{itemize}
  \item If $e=\istrf{s}$ then $e:\istr$.
  \item If $e=\ifilter{r}{\istrf{s}}$ then $e:\istr$.
  \item If $e=\istrf{s_1}+\istrf{s_2}$ then $e:\istr$.
\end{itemize}
\end{defn}

For simplicity, we assume a fixed translation from
$\lcs$ regular expressions to regular expressions recognizable by the internal
language's regex library (in practice, a fixed translation is acceptable.)
To summarize, we assume an internal language containing 
a string typetogether with operations for string concatentation and filtering. We
expect closure over strings for both operations.
Finally, recall that $\ifilter$ is only needed for dynamic casts, which may be removed without
descreasing the expressivity or even usability of the language.
Finally, the semantics of the filter function are defined in terms of ${\tt rl\_filter}$,
which is a static version of ${\tt ifilter}$.

\subsection{Properties of Regular Languages}

The regular languages are the smallest set generated by regular expressions
defined in Figure 1.

\begin{thm}{Closure Properties.} \label{thm:closure}
The regular expressions are closed under complements and concatenation.
\end{thm}
\begin{proof}
See \cite{cinderella}.
\end{proof}

\begin{thm}{Coercion Theorem.} \label{thm:coerce}
Suppose that $R$ and $L$ are regular expressions, and that $s \in R$ is a finite string.  Let $s' := coerce(R,L,s)$ with all maximal substrings recognized by $L$ replaced with $\epsilon$.  Then $s'$ is recognized by $(R \backslash L) + \emptyset$ and the construction of $R \backslash L$ is decidable.
\end{thm}
\begin{proof}
Let $F,G$ be FAs corresponding to $R$ and $L$, and let $G'$ be $G$ with its final states inverted (so that $G'$ is the complement of $L$).  Define an FA $H$ as a DFA corresponding to the NFA found by combining $F$ and $G'$ such that $H$ accepts only if $R$ and $L'$ accept or if $s$ is empty (this construction may result in an exponential blowup in state size.)  Clearly, $H$ corresponds to $R \backslash L + \emptyset$.  Thus, the construction of $R \backslash L + \emptyset$ is decidable.

If $R \subset L$, $s' = \emptyset$.  If $L \subset R$, either $s' = \emptyset$, or $s' \in R$ and $s' \not \in L$. If $R$ and $L$ are not subsets of one another, then it may be the case that $L$ recognizes part of $R$.  Consider $L$ as the union of two languages, one which is a subset of $R$ and one which is disjoint.  The subset language is considered above and the disjoint language is inconsequential.
\end{proof}

@@

@@

\begin{thm}[Preservation]
  Let $T$ be a type in $\lcs$ and $h(T)=\sigma$ the corresponding type in the internal language.
  For all terms $e$:
  \begin{itemize}
    \item If $e:T$ and $e:T \ireduces i$ then $i : \sigma$ such that $h(T) = \sigma$.
    \item If $e:T$ and $e:T \reduces e'$ then $e':T$.
  \end{itemize}
\end{thm}
\begin{proof}
The proof is a straightforward induction on the derivation of the combined evaluation relation.
\begin{itemize}[label=$ $,itemsep=1ex]
  \item \textbf{E-Ival, E-Ifilterval}. Both cases hold since the terms at hand are not $\lcs$ terms.
  \item \textbf{E-Stringval, E-strval}. Both cases hold since no reduction is possible. 
  \item \textbf{E-String}. By the definition of typing for internal terms, $\istrf{e} : \istr$. It suffices to show that $h(\str) = \istr$, which follows from the definition of $h$.
  \item \textbf{E-String\_in}. By the definition of typing for internal terms, $\istrf{e} : \istr$. It suffices to show that $h(\strin[r]) = \istr$, which follows from the definition of $h$ for arbitrary $r$.
  \item \textbf{E-concatval}. By the definition of typing for internal terms, $\istrf{e_1} + \istrf{e_2} : \istr$. So it suffices to show that $h(\strin[r_1+r_2]) = \istr$, which follows from the definition of $h$ for arbitrary $r_1$, $r_2$.
  \item \textbf{E-concatR, E-concatL}. WLOG consider E-concatR. By induction, $e_1' : \strin{l_1}$. By inversion of T-Concat at the premise, $e_2 : \strin{r_2}$. Therefore, $e_1 + e_2 : \strin[r_1 + r_2]$.
  \item \textbf{E-Filterval}. We have that $\filter{\strin[r']}{e}:\strin[r\backslash r' + \emptyset]$. By inversion of T-Filter, $e:\strin[r]$.
    By T-Equiv-string\_in (which is bidirectional), $e \in \lang{r}$.
    By the Coercion Theorem, ${\tt rl\_filter}(r,e) \in \lang{r\backslash r' + \emptyset}$.
    By T-Equiv-string\_in,  $e \in \lang{r}$ and ${\tt rl\_filter}(r,e) \in \lang{r\backslash r' + \emptyset}$ implies ${\tt rl\_filter}(r,e):\strin[r\backslash r' + \emptyset]$.
  \item \textbf{E-Filter}. By inversion, $e : \strin[r']$ $\reduces e'$ so by the induction $e':\strin[r']$.
    Therefore, \\${\tt filter}(\strin[r], e'):$ $\strin[r \backslash r' + \emptyset]$ by T-Filter.
  \item \textbf{E-Convertval}. It suffices to show that $h(\strin[r]) = \istr$, which is true by definition.
  \item \textbf{E-Convert}. By inversion and induction, $e' : \strin[r]$. We know that $[\strin[r']](e) : \strin[r']$, so by inversion of T-Convert $\lang{r'} \subseteq \lang{r}$.
    It follows that $[\strin[r']](e') : \strin[r']$.
  \item \textbf{E-DConvert}. By inversion, $e : \strin[r'] \reduces e'$. By the induction hypothesis, $e' : \strin[r']$; therefore, by T-Dconvert, $\dconvert[\strin[r]](e):\strin[r]$.
  \item \textbf{E-DConvertval}. By the definition of typing for internal terms, $\ifilter(r,\istrf{e}) : \istr$. It suffices to show that $h(\strin[r_1+r_2]) = \istr$, which follows from the definition of $h$.
\end{itemize}
\end{proof}

\begin{thm}[Progress]
  If $e:T$ then $e:T \reduces^* e' \ireduces^* i$ where $i \ival$.
\end{thm}
\begin{proof}
By induction on the derivation of $e:T$.
For \textbf{T-Equiv-Include}, note that $e \in \lang{r}$, $e = \strf{s}$ for some $s$; therefore, $e \val$ by E-Strinval. 
We have that $e \val$ and $e : \strin[r]$, so $e \ireduces \istrf{e}$ by E-string\_in.
The remaining cases follow by induction on the hypotheses and application of corresponding evaluation rules.
\end{proof}

@@

The $\lcs$ language is implemented as a type system extension in Ace; this extension
is illustrated in the examples at the beginning of this paper.

Computing a regular expression representating the language
$R \backslash S$ is necessary in order to type-check expressions in which the filter function
occurs. This language is computed by translating $R$ and $S$ into finite automata,
complementing the final states of $S$, then constructing the cross-product of $R$
and $S'$. Type checking terminates only because this construction is decidable.

In addition to this construction, other more typical operations -- such
as equality checks for regular expressions and regular expression matching -- are
also necessary. For these reasons, the Ace extension implementing $\lcs$ includes a library
implementing each of these constructions.
